Evolutionary rates of SARS-CoV-2

### 2022-08-07 manuscript/variant_rates.tex
edits

\begin{document}
\title{Evolutionary rates of SARS-CoV-2}
\author{Richard A.~Neher}
\date{\today}
\maketitle

@@

Due to shared ancestry, divergences of sequences are not independent data points and a regression against time is not suited to estimate evolutionary rates in general.
In particular, confidence intervals are difficult to obtain.
However, in the case of rapidly expanding variants we typically observe a large number of independent lineages emanating from one or several basal polytomies.
Along each of these lineages, mutation accumulation is independent and divergence increase allows to estimate the rate robustly.
Fig.~\ref{fig:rate_alpha} shows the increasing intra-clade divergence for clade 20I corresponding to VoC Alpha.
Both synonymous and non-synonymous within-clade average divergence increases linearly over time allowing for a robust estimate of the rate.

\begin{figure*}[tb]
    \includegraphics[width=\textwidth]{figures/rtt/20I_rtt.pdf}
    \caption{{\bf Divergence increases linearly with time (20I (Alpha)).} Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
    The blue line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter. Each panel also shows mean $\pm$ standard deviation and a weighted linear fit. Analogous figures for all clades considered are included in the appendix.
    \label{fig:rate_alpha}}
\end{figure*}

For almost all major Nextstrain clades, the average divergence increases linearly in time (similar patterns as Fig.~\ref{fig:rate_alpha}) allowing us to estimate clade specific evolutionary rates for amino acid and silent changes.
These rates are summarized in Fig.~\ref{fig:rate_summary} and Tab.~\ref{tab:rates}.

@@

@@

Nextstrain clades tend to be defined by long branches leading to a large polytomy.
It is thus expected that the estimated inter-clade rate exceeds the intra-clade rate purely because of this conditioning.
This effect might be particularly important early on in the pandemic where branches with as little two mutations were used to define new clades.
We thus also included composite clades 19B+, 19B++ and 20A+ containing sequences from 19A and 19B (rooted on 19B), 19A, 19B, 20A, 20B, 20C, 20D (rooted on 19B), and 20A, 20B, 20C, 20D (rooted on 20A).
The estimates for composite clades 19B+ and 20A+ are consistent with the estimates of the individual clades, while apparent rates of 19B++ are considerably higher.
The latter is due to the rapid expansion and subsequent dominance of clade 20A and its descendants \citep{korber_tracking_2020}.

\begin{table*}
\begin{tabular}{l|rrrrrr}
    \hline
    {\bf clade} &  overall rate $[y^{-1}]$ & aa rate $[y^{-1}]$ &  syn rate $[y^{-1}]$ & overall div. &  aa div. &  syn div. \\
    \hline
    19A &     19.76 &    12.79 &      6.97 &        0 &       0 &        0 \\
    19B &     19.47 &    12.37 &      7.10 &        2 &       1 &        1 \\
   19B+ &     18.55 &    11.90 &      6.65 &        2 &       1 &        1 \\
  19B++ &     28.21 &    16.37 &     11.84 &        2 &       1 &        1 \\
    20A &     15.66 &     8.40 &      7.25 &        4 &       2 &        2 \\
    20B &     14.14 &     7.61 &      6.54 &        7 &       4 &        3 \\
    20C &     16.30 &     9.52 &      6.78 &        6 &       4 &        2 \\
   20A+ &     17.15 &     9.81 &      7.33 &        4 &       2 &        2 \\
    20E &     13.65 &     5.74 &      7.91 &       11 &       4 &        7 \\
    20H &     14.50 &     8.03 &      6.47 &       19 &      15 &        4 \\
    20I &     12.18 &     5.85 &      6.33 &       28 &      18 &       10 \\
    20J &     13.31 &     6.84 &      6.47 &       33 &      21 &       12 \\
    21D &     11.36 &     6.29 &      5.07 &       25 &      12 &       13 \\
    21G &     12.05 &     6.52 &      5.53 &       26 &      18 &        8 \\
    21H &     18.17 &    10.00 &      8.17 &       29 &      20 &        9 \\
    21I &     16.33 &     8.67 &      7.66 &       28 &      23 &        5 \\
    21J &     12.25 &     6.32 &      5.93 &       32 &      27 &        5 \\
    21K &     14.37 &     6.70 &      7.67 &       54 &      45 &        9 \\
    21L &      9.02 &     3.80 &      5.22 &       66 &      49 &       17 \\
    22A &      8.10 &     2.79 &      5.31 &       69 &      51 &       18 \\
    22B &     12.07 &     4.97 &      7.11 &       64 &      49 &       15 \\
      \hline
    \end{tabular}
\caption{{\bf Evolutionary rates estimates from root-to-tip regressions for overall nucleotide changes, amino acid changes, and synonymous changes.}
    The last three columns give the distances of the clade founder sequence from the reference Wuhan/Hu-1.
\label{tab:rates}}
\end{table*}

@@

\subsection*{Diversification within clades}
A simple model for a growing variant is a super-critical branching process with growth rate $\alpha$ and an embedded mutation process.
Offspring of genomes with $i$ mutations contain $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
The probability that offspring genomes are identical to their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5days$ and a rate of $\mu = 15/year$ evaluates to $u=0.2$.
Since these branching process models are linear, them mean number of cases $n$ will increase exponentially with rate $\alpha$, while the number of genomes with $i$ mutations relative to the founder $m_i$ grows with rate $\alpha - u$ per generation.
\begin{equation}
    \begin{split}
        \frac{dn}{dt} &= \alpha n \\
        \frac{dm_0}{dt} &= (\alpha - u) m_0 \\
        \frac{dm_1}{dt} &= u m_0 (\alpha - u) m_1 \\
        \frac{dm_2}{dt} &= u m_1 (\alpha - u) m_2 \\
        \cdots
    \end{split}
\end{equation}
with solution $m_i = e^{(\alpha - u)t} \frac{(ut)^i}{i!}$.

For some clades, especially those that are well sampled soon after their emergence, this Poisson model is good fit to diversity accumulation and yields estimates of rates and time of origin that are compatible with the divergence regression, see Fig.~\ref{fig:within_clades}.
For other clades, especially when the founder sequence is barely observed and dominant subclades developed, such fits are not robust.
This is particularly true for the early clades 19A and 19B, as well as 21J (Delta), 21K (Omicron, BA.1), and 22B (Omicron, BA.5).

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/counts/20C_counts.pdf}
    \caption{{\bf Diversification within clades.} \label{fig:within_clades}}
\end{figure*}

These Poisson weights of mutation numbers are again only valid if mutations accumulate along many independent lineages.
In particular, this assumption is violated if some lineages spread systematically faster than others either because of epidemiological factors or the mutations they carry.

@@

When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and \ref{fig:fitness_landscape_supp}), the most constrained are ORF1b and M, while ORF3a, ORF6, ORF7a, ORF7b, and ORF8 show little evidence of constraint, consistent with frequently observed stop-mutations in these genes.
N shows an intermediate pattern, possibly reflecting its mix of structured and unstructured regions.
Rare variation at 3rd positions is very comparable between genes.
Only E shows slightly less variation at 3rd positions than other genes with a notable dip in the middle of the gene around codon 35.

@@

\section*{Discussion}
The inferred evolutionary rate of RNA viruses often decreases with the time scale across which is it estimated \citep{wertheim_purifying_2011,ghafari_prisoner_2021}.
This effect is particularly pronounced at the beginning of an outbreak following a host switch and has been attributed to methodological issues and incomplete purifying selection leading to inflated measures of diversity \citep{meyer_time_2015,ghafari_purifying_2022}.
In addition to segregating deleterious mutations, early viral evolution after a host switch will also be affect by anomalously fast adaptive evolution as a dramatic change in environment, i.e.~a host switch, likely results in many mutations that increase fitness, possibly only in epistatic combinations with others.
Such transient increases in the rate of adaptation is common in experimental evolution \citep{elena_virus_2007}.

The analysis presented here suggests that the early evolutionary rate of SARS-CoV-2 was inflated by adaptive evolution.
Since we analyzed evolution within short lived clades of SARS-CoV-2 over a span of 6 month, purifying selection has comparable efficiency for all clades.
Nevertheless, the non-synonymous evolutionary of clades circulating in late 2019 and early 2020 is about twice as high as that of clades in 2021 and 2020.

Adaptive evolution complicates phylodynamic analysis, which typically assumes that the mutation process is independent of the spread and epidemiology.
This assumption is true for neutral mutations that occur along every lineage with the same rate.
Adaptive and deleterious mutations, however, affect the ability of the virus to spread.

Deleterious mutations are relatively straightforward to account for: Strongly deleterious mutations don't spread and are only observed on terminal branches, similar to sequencing errors.
Weakly deleterious mutations can spread, but lineages that carry them tend to be short-lived.
A spectrum of mutations with different deleterious effects leads to time dependent effective evolutionary rates \citep{wertheim_purifying_2011}.
The intra-clade rates in Fig.~\ref{fig:rate_progression} are all estimated over a similar time frame and the time dependent rate phenomenon should not apply.

Adaptive evolution, however, is much harder to account for properly and can lead to more dramatic distortions.
Since the number of sites that allow beneficial mutations is small, adaptive evolution tends to be very stochastic.
Unlike neutral evolution, the rate of adaptive evolution depends on the population size and the transmission process.
Epistatic interactions can further steepen the dependence on population size and increase stochasticity.
Methods to infer emergence dates and locations of the early variants might therefore be biased or over-confident.

\bibliography{bib}
\appendix
\include{appendix}
\include{appendix_rtt}
\include{appendix_counts}
\end{document}

>>>>

\begin{document}
\title{Evolutionary rates of SARS-CoV-2}
\author{Richard A.~Neher}
\date{\today}

\begin{abstract}
The evolution of SARS-CoV-2 has had profound impact on the course of the pandemic and multiple variants of concern have led to renewed waves of infections.
Here we analyze the evolution of the virus over 2.5 years since its emergence and estimate rates of evolution for synonymous and non-synonymous changes within and between variants.
The rate of synonymous mutations is found to be around 6 changes per year.
Rates within variants vary little and are compatible with a global rate.
The non-synonymous changes within variants, however, was initially around 12-16 changes per year, but in 2021 and 2022 dropped to 6-9 changes per year.
Globally, the different variants of SARS-CoV-2 are compatible with an overall rate of about 25 amino acid changes per year, highlighting the qualitatively different adaptive evolutionary process gave rise to the different variants.
We further show that many accessory proteins evolve under limited evolutionary constraint with little short term purifying selection.
\end{abstract}

\maketitle

@@

Due to shared ancestry, divergences of sequences are not independent data points and a regression against time is not suited to estimate evolutionary rates in general.
In particular, confidence intervals are difficult to obtain.
However, in the case of rapidly expanding variants we typically observe a large number of independent lineages emanating from one or several basal polytomies.
Along each of these lineages, mutation accumulation is independent and divergence increase allows to estimate the rate robustly.
Fig.~\ref{fig:within_clade} shows the increasing intra-clade divergence for clade 20C, corresponding to a large clade that emerged in early 2020 that was common in North America and Scandinavia.
Both synonymous and non-synonymous within-clade average divergence increases linearly over time allowing for a robust estimate of the rate.

\begin{figure*}[tb]
    \includegraphics[width=\textwidth]{figures/rtt/20I_rtt.pdf}
    \includegraphics[width=\textwidth]{figures/counts/20C_counts.pdf}
    \caption{{\bf Within-clade divergence increases linearly with time.} Top: Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
    The green line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter. Each panel also shows mean $\pm$ standard deviation and a weighted linear fit. Analogous figures for all clades considered are included in the appendix.
    Bottom: Panels D and E show the prevalence of specific genotypes (D) and specific mutations (E). In the case shown, the founder genotype initially dominates and no daughter genotype or mutation dominate. Panel F shows a Poisson model to the breakdown of the population into genotypes with different number of mutations over time.
    \label{fig:within_clade}}
\end{figure*}

A simple model for diversity in a growing variant is a super-critical branching process with growth rate $\alpha$ and an embedded mutation process.
Offspring of genomes with $i$ mutations contain $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
The probability that offspring genomes are identical to their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5days$ and a rate of $\mu = 15/year$ evaluates to $u\approx 0.2$.

When considering a rapidly growing well sampled outbreak, typically a single founder genotype will give rise to a large number of daughter lineages that evolve independently.
In this case, the diversification processes is robustly described by its mean
Since the above branching process is linear, the mean number of cases $n$ will increase exponentially with rate $\alpha$, while the number of genomes with $i$ mutations relative to the founder $m_i$ grows with rate $\alpha - u$ per generation:
\begin{equation}
    \begin{split}
        \frac{dn}{dt} &= \alpha n \\
        \frac{dm_0}{dt} &= (\alpha - u) m_0 \\
        \frac{dm_1}{dt} &= u m_0 (\alpha - u) m_1 \\
        \frac{dm_2}{dt} &= u m_1 (\alpha - u) m_2 \\
        \cdots
    \end{split}
\end{equation}
with solution $m_i = e^{(\alpha - u)t} \frac{(ut)^i}{i!}$.
At each time the number of mutations in the population is expected to Poisson distributed with mean $ut$, while the overall number of cases is $e^{\alpha t}$ or more generally $e^{\int \alpha(t) dt}$ if growth rate or ascertainment varies over time.

For some clades, especially those that are well sampled soon after their emergence, this Poisson model is good fit to diversity accumulation and yields estimates of rates and time of origin that are compatible with the divergence regression, see Fig.~\ref{fig:within_clade}F for clade 20C.
Analogous plots for all other clades considered are included in the appendix.

In some cases, subclades do become dominant in the early stochastic dynamic.
This can result in increased within clade divergence, effectively generating an offset to diversity.
Extrapolation of the linear fit to zero divergence is thus not necessarily a good estimator of the emergence time of a variant.
For several clades 19B, 20H (Beta), 21D (Eta), 21G (Lambda), 21 (Mu), 21I (Delta), 21J (Delta), 21L (Omicron, BA.2), and 22B (Omicron, BA.5), founder like variants are a minority even in early data.

These Poisson weights of mutation numbers are again only valid if mutations accumulate along many independent lineages.
In particular, this assumption is violated if some lineages spread systematically faster than others either because of epidemiological factors or the mutations they carry.
In variant 21K (Omicron, BA.1), a sublineage with mutation $S:R346K$ might have enjoyed a transmission advantage.
A possible consequence of this advantage is visible in Supp.~Fig.~\ref{fig:21K_divergence} where genotypes with 0, 1, or 2 mutations are decaying more rapidly than those with more mutations.

@@

Despite these caveats, for almost all major Nextstrain clades, the average divergence increases linearly in time (similar patterns as Fig.~\ref{fig:within_clade}A-C for clade 20C) allowing us to estimate clade specific evolutionary rates for amino acid and silent changes.
These rates are summarized in Fig.~\ref{fig:rate_summary} and Tab.~\ref{tab:rates}.

@@

Nextstrain clades tend to be defined by long branches leading to a large polytomy.
It is thus expected that the estimated inter-clade rate exceeds the intra-clade rate purely because of this conditioning.
This effect might be particularly important early on in the pandemic where branches with as little two mutations were used to define new clades.
We thus also included composite clades 19B+, 19B++ and 20A+ containing sequences from 19A and 19B (rooted on 19B), 19A, 19B, 20A, 20B, 20C, 20D (rooted on 19B), and 20A, 20B, 20C, 20D (rooted on 20A).
The estimates for composite clades 19B+ and 20A+ are consistent with the estimates of the individual clades, while apparent rates of 19B++ are considerably higher.
The latter is due to the rapid expansion and subsequent dominance of clade 20A and its descendants which rapidly added four additional mutations \citep{korber_tracking_2020}.

\begin{table*}
\begin{tabular}{l|rrrrrr}
    \hline
    {\bf clade} &  overall rate $[y^{-1}]$ & aa rate $[y^{-1}]$ &  syn rate $[y^{-1}]$ & overall div. &  aa div. &  syn div. \\
    \hline
    19B &     19.45 &    12.37 &      7.09 &        0 &       0 &        0 \\
    19B+ &     18.51 &    11.87 &      6.64 &        0 &       0 &        0 \\
    19B++ &     28.17 &    16.32 &     11.84 &        0 &       0 &        0 \\
    19A &     19.75 &    12.79 &      6.96 &        2 &       1 &        1 \\
    20A &     15.59 &     8.33 &      7.26 &        6 &       3 &        3 \\
     20B &     14.17 &     7.64 &      6.53 &        9 &       5 &        4 \\
     20C &     16.43 &     9.63 &      6.79 &        8 &       5 &        3 \\
    20A+ &     17.16 &     9.76 &      7.40 &        6 &       3 &        3 \\
     20E &     13.88 &     5.84 &      8.04 &       13 &       5 &        8 \\
     20H &     15.46 &     9.50 &      5.96 &       21 &      16 &        5 \\
     20I &     11.78 &     5.72 &      6.06 &       30 &      19 &       11 \\
     20J &     12.91 &     6.29 &      6.62 &       35 &      22 &       13 \\
     21D &     11.36 &     6.29 &      5.07 &       27 &      13 &       14 \\
     21G &     12.03 &     6.51 &      5.51 &       28 &      19 &        9 \\
     21H &     18.24 &    10.05 &      8.19 &       31 &      21 &       10 \\
     21I &     16.26 &     8.63 &      7.63 &       30 &      24 &        6 \\
     21J &     13.26 &     6.91 &      6.36 &       34 &      28 &        6 \\
     21K &     14.04 &     6.55 &      7.49 &       56 &      46 &       10 \\
     21L &      9.40 &     4.12 &      5.28 &       68 &      50 &       18 \\
     22A &      8.07 &     2.84 &      5.23 &       71 &      52 &       19 \\
     22B &     12.06 &     4.96 &      7.10 &       66 &      50 &       16 \\
         \hline
    \end{tabular}
\caption{{\bf Evolutionary rates estimates from root-to-tip regressions for overall nucleotide changes, amino acid changes, and synonymous changes.}
    The last three columns give the distances of the clade founder sequence from putative MRCA of SARS-CoV-2.
\label{tab:rates}}
\end{table*}

@@

@@

When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and \ref{fig:fitness_landscape_supp}), the most constrained are \emph{ORF1b} and \emph{M}, while \emph{ORF3a, ORF6, ORF7a, ORF7b}, and \emph{ORF8} show little evidence of constraint, consistent with frequently observed stop-mutations in these genes.
\emph{N} shows an intermediate pattern, possibly reflecting its mix of structured and unstructured regions.
Rare variation at 3rd positions is very comparable between genes.
Only \emph{E} shows slightly less variation at 3rd positions than other genes with a notable dip in the middle of the gene around codon 35.

@@

\section*{Discussion}
The inferred evolutionary rate of RNA viruses often decreases with the time scale across which is it estimated \citep{wertheim_purifying_2011,ghafari_prisoner_2021}.
This effect is particularly pronounced at the beginning of an outbreak following a host switch and has been attributed to methodological issues and incomplete purifying selection leading to inflated measures of diversity \citep{meyer_time_2015,ghafari_purifying_2022}.
In addition to segregating deleterious mutations, early viral evolution after a host switch will also be affected by anomalously fast adaptive evolution as a dramatic change in environment, i.e.~a host switch, likely results in many mutations that increase fitness, though sometimes these better adapted states require multiple epistatically interacting mutations \citep{martin_emergence_2021}.
Such transient increases in the rate of adaptation is common in experimental evolution \citep{elena_virus_2007}.

Since we analyzed evolution within short lived clades of SARS-CoV-2 over a span of 6 month, purifying selection has comparable efficiency for all clades.
Sequencing artefacts are expected to increase divergence at all time points or decrease with time as sequencing workflows improve.
Nevertheless, the estimated non-synonymous evolutionary rate of clades circulating in late 2019 and early 2020 is about twice as high as that of clades in 2021 and 2022, while the synonymous rate does not change (see Fig.~\ref{fig:rate_summary} and Table \ref{tab:rates}).
One possible explanation is that the early evolutionary rate of SARS-CoV-2 was inflated by adaptive evolution.

Most of this apparent acceleration of within-variant evolution ceased in mid 2020 and even early variants like 20E accumulated non-synonymous changes at a rate of about 6 per year instead of 12.
By that time, the number of non-synonymous mutations was SARS-CoV-2 small (5 in the case of 20E) and given what is observed in subsequent variants, it is implausible that this small change would have exhausted the pool of beneficial mutations.
However, the different replication and transmission kinetics of variants with S:D614G might mean that virus is operating closer to the maximal capacity at which cells can produce virions and the strength of selection is generally reduced.
Difference in diversification of SARS-CoV-2 with and without the S:D614G mutation where also observed in evolution experiments \citep{amicone_mutation_2022}, who concluded that these differences were not due to a change in base line mutation rate.

An evolutionary rate of 6 synonymous changes at around 9000 positions per year corresponds to a per-site mutation rate of $6.6\times 10^{-4}$ per site and year or $1.8\times 10^{-6}$/day, slightly higher than the estimated base-line mutation rate of $1.3\times 10^{-6}$/day \citep{amicone_mutation_2022}.
The overall rate in variants starting with 20E is in the range of 9-16 changes per year, corresponding to a per site rate of $3 - 5\times 10^{-4}$/site and year, consistent with recent estimates by \citet{hill_origins_2022} and \citet{tay_emergence_2022}.

When considering only the putative founder genotype and date of origin of each variant, all variants so far are compatible with a back-bone evolutionary rate of 32 changes per year, corresponding to an per site rate of around $10^{-3}$/site/year (see Fig.~\ref{fig:rate_summary}).
This rate is a composite of episode of cryptic accelerated evolution (probably in chronically infected individuals), and regular transmission chains of acute infections.
This is thus lower than the branch specific rates estimated by \citet{hill_origins_2022} and \citet{tay_emergence_2022}.
However, all clades and variants are compatible with a single ``back-bone'' molecular clock without postulating significant differences in the emergence of different variants.
This observation of course does not rule out that different clades and variants emerged in qualitatively different ways.

By analysing the rate mutations that spread only on short times scales within fine-grained Pango lineages, we estimated the level of constraint on different parts of the SARS-CoV-genome.
The great majority of 3rd positions in codons at which  most mutations are synonymous do not show strong signatures of conservations.
The major open reading frames ORF1ab, S, N, E, M show clear signatures of purifying selection with around 50\% of sites being more constrained than the most constrained 10\% of 3rd positions.
This result is consistent with similar analysis in HIV that also show little fitness costs for most synonymous mutations, while half of non-synonymous mutations are so deleterious that they are not even observed on short times scales \citep{zanini_vivo_2017}.
The remaining ORFs show little global signal of constraint and 1st and 2nd positions are as variable as 3rd positions (see Fig.~\ref{fig:fitness_costs_supp}).
Only a few regions show a clear signal of conservation at 3rd positions, notably a central region of \emph{E} and the ribosome slippage region at the beginning of \emph{ORF1b}.

The heterogeneity in evolutionary rates and the combination of adaptive evolution, approximately neutral mutations, and purifying selection complicate the interpretation of phylodynamic analysis, which typically assumes that the mutation process is independent of the spread and epidemiology and independent at differents.
These assumptions are (approximately) true for neutral mutations that occur along every lineage with the same rate.
Deleterious mutations are relatively straightforward to account for: Strongly deleterious mutations don't spread and are only observed on terminal branches, similar to sequencing errors.
Weakly deleterious mutations can spread, but lineages that carry them tend to be short-lived.
A spectrum of mutations with different deleterious effects leads to time dependent effective evolutionary rates \citep{wertheim_purifying_2011}.

Adaptive evolution, however, is much harder to account for properly and can lead to more dramatic distortions.
Since the number of sites that allow beneficial mutations is small, adaptive evolution tends to be very stochastic.
Unlike neutral evolution, the rate of adaptive evolution depends on the population size and the transmission process.
Methods to infer emergence dates and locations of the early variants might therefore be biased or over-confident.

\bibliography{bib}
\appendix
\include{appendix}
\onecolumngrid
\include{appendix_rtt}
% \include{appendix_counts}
\end{document}
